---
title: "GOexpress: identify and visualise robust gene ontology signatures
    through supervised classification of gene expression data"
author:
    -   name: Kévin C. Rue-Albrecht
        email: kevinrue67@gmail.com
        affiliation:
            -   University College Dublin
            -   University of Oxford
    -   name: Paul A. McGettigan
        affiliation: University College Dublin
    -   name: Belinda Hernández
        affiliation: University College Dublin
    -   name: David A. Magee
        affiliation: University College Dublin
    -   name: Nicolas C. Nalpas
        affiliation: University College Dublin
    -   name: Andrew C. Parnell
        affiliation: University College Dublin
    -   name: Stephen V. Gordon
        affiliation: University College Dublin
    -   name: David E. MacHugh
        affiliation: University College Dublin
package: GOexpress
output: BiocStyle::html_document2
abstract: |
    Introduction to the concepts implemented in the package.
vignette: |
    %\VignetteIndexEntry{Introduction}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
bibliography:
    bibliography.bib
---

```{r GOexpress, include=FALSE}
stopifnot(require('GOexpress'))
```

# Introduction

## The origin and purpose of GOexpress

`r Biocpkg("GOexpress")` emerged from a set of plotting
functions regularly copied and pasted across various complex multifactorial
transcriptomics studies from both microarray and RNA-seq platforms.
Those functions were repeatedly used to:

* visualise and compare the expression profile of genes
    across multiple groups of samples
* annotate technical gene identifiers from both microarray and RNA-seq
    platforms (*i.e.*, probesets, Ensembl gene identifiers)
    with their associated gene name,
* evaluate the classification of samples based on genes participating
    in a common cellular function or location (*i.e.*, gene ontology).
    
While developing `r Biocpkg("GOexpress")` and discussing its features with
colleagues and potential users, a few more features were added,
to enhance and complement the initial functions,
leading to the present version of the package.

## Purpose of this document

This User’s Guide was intended as a helpful description of the main features
implemented in `r Biocpkg("GOexpress")`, as well as a tutorial taking the user
through a typical analysis pipeline that `r Biocpkg("GOexpress")` was designed
for.
While an example usage will be provided for each function of the package,
the many arguments of each function cannot realistically be demonstrated in
this Guide, and we kindly ask users to also read the individual help pages
accompanying the corresponding package functions for further details.

## Versions 2.x.x

Recently, `r Biocpkg("GOexpress")` has undergone important *structural*
changes to introduce S4 classes and methods.
Crucially, every effort has been made to preserve the underlying algorithms
and functionalities, and minimise disruption of the user experience.
Ultimately, those changes are intended to facilitate the user experience,
and widen the range of data types supported.

Among the most notable changes:

* The `GO_analyse` function implemented in versions `<=1.x.x` is now split
    into smaller S4 methods with specific signatures and output classes.
    * `oneway.test` replaces `GO_analyse(...,method='anova)`'
    * `randomForest` replaces `GO_analyse(...,method='randomForest')`
    * `GOrank` replaces the portion of `GO_analyse` that used to summarise
        features annotated to GO categories.
* Semi-automated support for the Ensembl BioMart was removed. Users
    are encouraged to query annotations directly
    (*e.g.*, `r Biocpkg("biomaRt")`).
    * However, the `GOMap` class was introduced to help users store annotations
        and track information about their source.
        Note that the `GOMap` class may be used to store *any* kind of
        bidirectional annotations between features and categories
        that users wish to use
        (*e.g.*, protein, miRNA, KEGG pathways, ...)

# Before you start

## Installation

Installing GOexpress should be as easy as running the two lines below:

```{r biocLite, eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite("GOexpress")
```

## Getting help

Despite our efforts to repeatedly test the package on various data sets,
many of the model species and gene expression platforms have
not been tested yet. We welcome feedback!

Interesting suggestions for additional package functions, or improvement of
existing ones are most welcome and may be implemented as time allows.
Alternatively, we also encourage users to fork the GitHub repository of the
project, develop and test their own feature(s), and finally generate a pull
request to integrate it to the original
[repository](https://github.com/kevinrue/GOexpress).

As for all \Bioconductor{} packages, the official
[support site](http://www.bioconductor.org/help/support)
is the best place to seek advice from a large and active community of
users and developers.

## Citing GOexpress

If you use `r Biocpkg("GOexpress")` in your work, please cite the associated
publication [@GOexpress2016]:

```{r citation, results='asis', echo=FALSE}
cat(citation('GOexpress')$textVersion)
```

# Input data

## ExpressionSet

`r Biocpkg("GOexpress")` requires an expression data matrix
and phenotype information in `assayData` and `phenoData` of 
a `r Biocpkg("Biobase")` `ExpressionSet`, for two main reasons:

* built-in methods readily available to store, access, update, and process the
    data set
* interoperatibility with other packages that produce and
    accept `ExpressionSet` objects

The other slots of the `ExpressionSet` container may be left empty as
`r Biocpkg("GOexpress")` does not currently access them.
Instructions to create `AnnotatedDataFrame` and `ExpressionSet`
objects are detailed in the vignettes of the `r Biocpkg("Biobase")` package.

### Expression data matrix

Despite their different underlying technologies, omics data sets
typically yield a matrix measuring the expression level of
numerous features in several samples.
Commonly, this expression matrix is filtered to
retain only genes qualified as *informative* (*e.g.*, above a detection level);
and uninformative features are removed to limit the false discovery rate (FDR).
`r Biocpkg("GOexpress")` uses the expression matrix supplied *as is*;
therefore, users should ensure that adequate filtering and normalisation
were applied before using dedicated packages (*e.g.*, `edgeR`, `DESeq2`).

In this vignette, we use a subset of a gene-by-sample matrix normalised
log[2]-transformed CPM from an *in vitro* stimulation of bovine macrophages with
different mycobacterial strains:

```{r Bt.logCPM, results='hold'}
data("Bt.logCPM")
class(Bt.logCPM); dim(Bt.logCPM)
```

**Note:**
See `?Bt.logCPM` for more information.

### Phenotype information

`r Biocpkg("GOexpress")` requires the expression data matrix
to be accompanied by a `r Biocpkg("Biobase")` `AnnotatedDataFrame`
that provides phenotype information for each sample
(*e.g.*, treatment, time point):

```{r Bt.pheno}
data("Bt.pheno")
summary(Bt.pheno)
```

**Note:**
To use `r Biocpkg("GOexpress")` in the best conditions,
users should ensure that experimental factors and levels are correctly set
(*e.g.*, drop unused levels).

```{r droplevels}
Bt.pheno <- droplevels(Bt.pheno)
```

### Feature information

Feature annotations may be useful to [subset]() the `ExpressionSet`,
or control elements of subsequent plots (*e.g.*, facets, aesthetics, labels).

**Note**
Currently, the `ExpressionSet` included for demonstration does not
include any annotations in the `featureData` slot. Therefore, let us
create a dummy annotation for demonstration purposes:

```{r dummy}
Bt.feature <- data.frame(
    dummy = sample(LETTERS, nrow(Bt.logCPM), TRUE),
    row.names = rownames(Bt.logCPM),
    stringsAsFactors = FALSE
)
```


## Experimental factor analysed

The above `ExpressionSet` and the name of a column in its `phenoData` slot
are the only two input variables required to start an analysis of the
expressed features.

In the sample data set, the next chunk of code reveals which columns store
factors:

```{r pDataClass}
sapply(Bt.pheno, class)
```

## Functional annotations

To summarise information associated with functionally
related or similar groups of expressed features,
`r Biocpkg("GOexpress")` allows the use of annotations such as the
Gene Ontology (GO) (*i.e.*,
*molecular function*, *biological process*, and *cellular component*).

`r Biocpkg("GOexpress")` implements the `GOMap` class to store any type
of bidirectional annotations between features and categories, along with
some information about their origin:

```{r Bt.GOMap, results='hold'}
data("Bt.GOMap")
Bt.GOMap
```

# Quick start

## ExpressionSet

A `matrix` of expression data and a `data.frame` of phenotype information
may be combined into an `ExpressionSet` as follows:

```{r ExpressionSet}
Bt.eset <- ExpressionSet(
    assayData = Bt.logCPM,
    phenoData = AnnotatedDataFrame(Bt.pheno),
    featureData = AnnotatedDataFrame(Bt.feature)
)
```

## Random Forest

Here, we demonstrate the use of a random forest (RF) approach to estimate
the capacity of each gene feature to classify samples subjected to different
infections:

```{r randomForest, results='hold'}
rf <- randomForest(Bt.eset, 'Infection')
```

The most interesting features for supervised classification may be extracted
from the `randpomForest` object as follows:

```{r topFeatures}
topFeatures(rf, n = 3)
```

## Rank categories

Here, we demonstrate how the average rank of features annotated to each
category can be calculated:

```{r GOrank}
data("Bt.GOMap")
gr <- GOrank(rf, Bt.GOMap)
gr
```

**Note:**
This analysis considers the `Infection` factor *only*,
irrespective of other experimental factors such as `Timepoint` and `Animal`;
to reach a high statistic, features must show a consistent difference between
`Infection`, independently of other factors.
Alternatively, a subset of samples from the input `ExpressionSet`
may be supplied to address more specific hypotheses.

The `randomForest` returns a list of class `randomForest`
that benefits from all the methods implemented in the original package:

```{r plotRandomForest}
plot(rf)
```

# Advanced use

## Subsetting ExpressionSet objects

`r Biocpkg("GOexpress")` implements a `subset` S4 method for `ExpressionSet`
objects. This method facilitates analyses of subsets of samples and features
using logical expressions applied to the `phenoData` and `featureData` slots:

```{r subset}
Bt.sub <- subset(Bt.eset, pheno = (Timepoint == '2H'), feature = (dummy < 'N'))
dim(Bt.eset); dim(Bt.sub)
```

**Note:**
The `subset` method returns an `ExpressionSet` object that can be directly
passed to an analysic or plotting method.

# Visualisation

`r Biocpkg("GOexpress")` implements several methods to visualise and share
expression profiles of interest. In versions **>=2.x.x**, `ggplot2` was
established as the primary plotting system for `r Biocpkg("GOexpress")`.

```{r ggplot2}
require(ggplot2)
```

For this purpose, let us identify:

* the most important feature identified by the `randomForest` method:

```{r bestFeature.rf}
bestFeature.rf <- rownames(topFeatures(rf, 1))
```

* the least important feature identified by the `randomForest` method:

```{r worstFeature.rf}
worstFeature.rf <- tail(rownames(topFeatures(rf, nrow(rf$importance))), 1)
```

**Note:**
For the `randomForest`, features best classifying a particular group of samples
may be identified using the corresponding column in `importance(rf)`. 
For instance, `topFeatures(rf,value="CN")`.

## Smoothed feature profile

We may plot the smoothed expression profile of the most important feature,
indicating each `Infection` group by a different colour:

```{r smoothFeature_feature, message=FALSE, warning=FALSE}
smoothFeature(Bt.eset[bestFeature.rf,], x = "Time", group = "Infection")
```

It is possible to visualised simultaneously multiple features;
in that case, the `smoothFeature` creates a new facet for each feature:

```{r smoothFeature_features, message=FALSE, warning=FALSE}
smoothFeature(
    Bt.eset[c(bestFeature.rf, worstFeature.rf),],
    x = "Time", group = "Infection"
) +
    theme_bw() +
    labs(
        y = "Expression",
        title = "Most and least important features identified by randomForest"
    )
```

**Note:**
The `smoothFeature` method returns an `ggplot` object that can be further
configured (*e.g.*, `theme`, `labs`), as shown above.

## Combine assay, feature, and phenotype data for complete freedom

The `ggfy` method combines expression data for any number of features
with fields from the `phenoData` and `featureData` slots
in a `data.frame` ready for use by `r CRANpkg("ggplot2")` methods:

```{r ggfy}
ggplot(ggfy(
    Bt.eset[c(bestFeature.rf, worstFeature.rf),],
    pheno = c("Animal", "Time", "Infection"),
    feature = c("dummy"),
    assay = "exprs"
)) +
    facet_grid(feature ~ Animal) +
    geom_line(aes(Time, exprs, colour = Infection, linetype = dummy)) +
    scale_y_continuous(limits = range(exprs(Bt.eset))) +
    theme_bw()
```


# Notes

## Importance of feature annotations

During the summarisation of feature into categories, features that appear in
the annotations but are *absent from the data set* are considered to have
no classification power, and therefore ranked last (*i.e.*, count of features
present in the data set plus one).

In data sets where uninformative features were filtered out prior to
the analysis, this ensure that categories under-represented in the data set
(*i.e.*, contain few informative features) are weighed down when considering
the average rank of annotated features.
Conversely, this prioritises categories that are well-represented in the data
set (*i.e.*, contain mostly informative features).

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References
